---
title: "Bulk RNA-seq Pathway Scoring Notebook"
output: html_notebook
---

This notebook runs the same pathway-scoring pipeline as `pathway_scoring.R`, but in an RStudio-friendly format.

How to use it:

1. Open this `.Rmd` file in RStudio.
2. Put your curated gene-set CSV/TSV/GMT in this folder, or fill in the starter `curated_gene_sets.csv` template.
3. Edit the `geneset_file` line in the configuration chunk only if you want to point to a different filename.
4. Run the notebook top to bottom, or click **Knit**.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)

analysis_dir <- normalizePath(".", winslash = "/")
source(file.path(analysis_dir, "pathway_scoring.R"), local = globalenv())
```

## Configuration

Update the file paths or parameters here as needed. The defaults below match the DESeq2-style avian bulk RNA-seq table already present in this folder.

```{r configuration}
input_file <- "DESeq2_full_combined_ssRNAseq.csv"
geneset_file <- "curated_gene_sets_populated.csv"   # Change only if your curated CSV/TSV/GMT has a different filename.
outdir <- "pathway_scoring_results_notebook"

methods <- c("gsva", "ssgsea", "singscore")
gene_id_col <- "SYMBOL"
expr_prefix <- "norm_"
group_pattern <- "^norm_([A-Za-z]+)_([0-9]+)$"

min_set_size <- 5L
max_set_size <- 500L
low_expr_threshold <- 1
low_expr_min_samples <- 2L
disable_low_expr_filter <- FALSE

stats_test <- "kruskal"
pairwise_test <- "none"      # Use "auto", "wilcox", or "t" to calculate pairwise comparisons.
pairwise_correction <- TRUE  # Set FALSE to report uncorrected pairwise p-values.
adjust_method <- "BH"
plot_method_correlation <- FALSE
boxplot_significance <- TRUE
```

```{r preflight}
starter_pathways <- c(
  "necroptosis",
  "TNF_NFkB",
  "TLR_cGAS_STING",
  "complement",
  "heterophil_myeloid",
  "angiogenesis",
  "osteogenesis",
  "ECM_remodeling",
  "sterol_cholesterol_fatty_acid_metabolism",
  "fracture_healing"
)

write_starter_geneset_template <- function(path) {
  starter_tbl <- tibble::tibble(
    set_name = starter_pathways,
    gene_symbol = "",
    direction = "",
    description = ""
  )
  readr::write_csv(starter_tbl, path, na = "")
}

if (!file.exists(input_file)) {
  stop("Input file not found: ", input_file)
}

if (!file.exists(geneset_file)) {
  candidate_gene_sets <- list.files(
    path = analysis_dir,
    pattern = "\\.(csv|tsv|txt|gmt)$",
    full.names = FALSE,
    ignore.case = TRUE
  )
  candidate_gene_sets <- setdiff(candidate_gene_sets, basename(input_file))

  if (length(candidate_gene_sets) == 1) {
    geneset_file <- candidate_gene_sets[[1]]
    message("Auto-detected gene-set file: ", geneset_file)
  } else if (length(candidate_gene_sets) > 1) {
    stop(
      "Multiple candidate gene-set files were found: ",
      paste(candidate_gene_sets, collapse = ", "),
      "\nSet 'geneset_file' explicitly in the configuration chunk."
    )
  } else {
    write_starter_geneset_template(geneset_file)
    stop(
      "No gene-set file was found, so a starter template was created at: ", geneset_file,
      "\nFill in the gene_symbol column with your curated pathway genes, then rerun the notebook."
    )
  }
}

if (tolower(tools::file_ext(geneset_file)) != "gmt") {
  delim <- if (grepl("\\.(tsv|tab|txt)$", tolower(geneset_file))) "\t" else ","
  reader <- if (identical(delim, "\t")) readr::read_tsv else readr::read_csv
  gene_set_preview <- reader(geneset_file, show_col_types = FALSE, progress = FALSE)
  set_name_col <- find_ci_col(names(gene_set_preview), "set_name")
  gene_symbol_col <- find_ci_col(names(gene_set_preview), "gene_symbol")

  if (is.na(set_name_col) || is.na(gene_symbol_col)) {
    stop(
      "Gene-set file must contain columns named set_name and gene_symbol.\n",
      "Found columns: ", paste(names(gene_set_preview), collapse = ", ")
    )
  }

  gene_symbols <- trimws(as.character(gene_set_preview[[gene_symbol_col]]))
  if (length(gene_symbols) == 0 || all(!nzchar(gene_symbols) | is.na(gene_symbols))) {
    stop(
      "The gene-set file exists but the gene_symbol column is still blank.\n",
      "Either populate ", geneset_file, " with your curated pathway genes,\n",
      "or change 'geneset_file' in the configuration chunk to point to your real curated CSV/TSV/GMT file."
    )
  }
}

if (!file.exists(geneset_file)) {
  stop(
    "Gene-set file not found: ", geneset_file,
    "\nUpdate 'geneset_file' in the configuration chunk to point to your curated CSV/TSV/GMT file."
  )
}

if ("singscore" %in% methods && !requireNamespace("singscore", quietly = TRUE)) {
  message(
    "The notebook is configured to run singscore, but the 'singscore' package is not installed.\n",
    "Install it first, or temporarily set methods <- c(\"gsva\", \"ssgsea\")."
  )
}
```

## Run Pathway Scoring

```{r run-pathway-scoring}
opts <- build_pathway_scoring_options(
  input = input_file,
  genesets = geneset_file,
  outdir = outdir,
  methods = methods,
  gene_id_col = gene_id_col,
  expr_prefix = expr_prefix,
  group_pattern = group_pattern,
  min_set_size = min_set_size,
  max_set_size = max_set_size,
  low_expr_threshold = low_expr_threshold,
  low_expr_min_samples = low_expr_min_samples,
  disable_low_expr_filter = disable_low_expr_filter,
  stats_test = stats_test,
  pairwise_test = pairwise_test,
  pairwise_correction = pairwise_correction,
  adjust_method = adjust_method,
  plot_method_correlation = plot_method_correlation,
  boxplot_significance = boxplot_significance
)

results <- run_pathway_scoring(opts)
```

## Output Summary

```{r output-summary}
results$output_dirs
```

```{r output-files}
unlist(results$output_paths, recursive = TRUE, use.names = TRUE)
```

## Sample Metadata

```{r sample-metadata}
knitr::kable(results$sample_metadata)
```

## Gene-set Coverage

```{r gene-set-coverage}
knitr::kable(results$coverage)
```

## Statistics Preview

```{r statistics-preview}
stats_preview <- purrr::imap_dfr(
  results$stats,
  ~ dplyr::mutate(.x, method = .y),
  .id = NULL
)

knitr::kable(utils::head(stats_preview, 15))
```

## Plot Files

The pipeline writes the heatmaps, boxplots, and PCA PDFs into the `plots/` subfolder inside `outdir`.

```{r plot-files}
list.files(file.path(results$output_dirs$outdir, "plots"), full.names = TRUE)
```
