---
title: "DESeq2_Clemson_output GOS refactor– UPDATED (2025‑06‑30)"
author: "Rooksie Noorai (original)--> Galen O'Shea‑Stone (refactor 6/25)"
date: "2025‑06‑30"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(0666)              # reproducibility for any resampling/IHW
```

## 1  Load libraries
```{r libs}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

#BiocManager::install(c(
 # "DESeq2", "apeglm", "clusterProfiler", "enrichplot",
  #"org.Gg.eg.db", "AnnotationDbi", "pathview"), ask = FALSE)     #use these as needed

library(DESeq2)
library(apeglm)              # for LFC shrinkage
library(clusterProfiler)
library(enrichplot)
library(AnnotationDbi)
library(org.Gg.eg.db)
library(pathview)
library(ggplot2)
```

## 2  Read counts & metadata
```{r read-counts}
# FULL set (commented out)
# Expression <- read.delim("redo.ensembl.stranded.counts.renamed.txt")
# SUBSET used below ──────────────────────────────────────────────────────────
Expression <- read.delim("redo.ensembl.stranded.counts.renamed.subset.txt")
rownames(Expression) <- Expression$Geneid

# keep only raw counts (remove Geneid column)
cts <- as.matrix(Expression[ , -1])
storage.mode(cts) <- "integer"

# condition & (constant) library type
condition <- factor(c("controlF","controlF","experimental","controlP",
                      "experimental","experimental","controlP",
                      "controlF","controlP","controlP"),
                    levels = c("controlF", "controlP", "experimental"))
coldata   <- data.frame(condition = condition, row.names = colnames(cts))
```

## 3  Create DESeqDataSet & pre-filter
```{r dds-setup}
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData   = coldata,
                              design    = ~ condition)

# keep genes with ≥10 counts in ≥3 samples (DESeq2 guideline)
keep <- rowSums(counts(dds) >= 10) >= 3
dds  <- dds[keep, ]
```

## 4  Run DESeq2
```{r deseq-run}
dds <- DESeq(dds)
resultsNames(dds)            # view available contrasts
```

## 5  Log-free variance stabilisation (VST) & QC plots
```{r vst-qc, fig.width = 6.5, fig.height = 5}
vsd <- vst(dds, blind = FALSE)

# PCA
p <- plotPCA(vsd, intgroup = "condition") + ggtitle("PCA – variance-stabilised")
print(p)

ggsave("DESeq2_PCA_plot.pdf", p, width = 6, height = 5)

# size-factor barplot
barplot(sizeFactors(dds), main = "Normalisation size factors",
        las = 2, ylab = "sizeFactor")

# mean–dispersion trend
plotDispEsts(dds)
```

## 6  Primary contrast: experimental vs controlF
```{r primary-contrast}
# Wald statistics (raw, unshrunk)
res_raw <- results(dds, name = "condition_experimental_vs_controlF")

# Shrink LFC for publication & GSEA ranking
res_shr <- lfcShrink(dds, coef = "condition_experimental_vs_controlF",
                     type = "apeglm")

# Merge: keep raw p‑values/FDR but use shrunken effect sizes
res_final <- data.frame(
  baseMean        = res_raw$baseMean,
  log2FC_shrunken = res_shr$log2FoldChange,
  lfcSE_shrunken  = res_shr$lfcSE,
  pvalue          = res_raw$pvalue,
  padj            = res_raw$padj,
  row.names       = rownames(res_raw)
)

write.csv(res_final, "DESeq2_experimental_vs_controlF.csv")
```

## 7  Secondary contrast: experimental vs controlP  Secondary contrast: experimental vs controlP
```{r secondary-contrast}
# re-level reference
dds_P <- dds
dds_P$condition <- relevel(dds_P$condition, ref = "controlP")
dds_P <- DESeq(dds_P)

res_P_shr <- lfcShrink(dds_P, coef = "condition_experimental_vs_controlP",
                       type = "apeglm")
core_cols_P <- intersect(c("baseMean","log2FoldChange","lfcSE","stat",
                           "pvalue","padj","svalue"), colnames(res_P_shr))
write.csv(as.data.frame(res_P_shr)[ , core_cols_P],
          "DESeq2_experimental_vs_controlP.csv")
```

## 8  Export normalised counts for reviewers
```{r export-counts}
write.csv(counts(dds, normalized = TRUE), "normalised_counts.csv")
```
#####Two chunks here from original script for doing a gsea go/KEGG analysis and some basic visualization, uncomment to run/revise (does not run as of 6/30/25)

## 9  Gene-set enrichment (clusterProfiler) – GO
#```{r gsea-go}
# Prepare ranked gene list (all genes)
ranked <- res_shr$log2FoldChange
names(ranked) <- rownames(res_shr)               # ENSEMBL IDs already
ranked <- sort(na.omit(ranked), decreasing = TRUE)

# GSEA (BH adjusted)
gse_go <- gseGO(geneList     = ranked,
                keyType      = "ENSEMBL",
                ont          = "ALL",
                OrgDb        = org.Gg.eg.db,
                nPerm        = 10000,
                minGSSize    = 3,
                maxGSSize    = 800,
                pvalueCutoff = 0.05,
                pAdjustMethod = "BH",
                verbose      = TRUE)

pdf("GO_GSEA_dotplot.pdf")
print(dotplot(gse_go, showCategory = 10, split = ".sign") +
        facet_grid(. ~ .sign))
dev.off()
#```

## 10  Gene-set enrichment – KEGG
#```{r gsea-kegg}
# Convert ENSEMBL → ENTREZ
ids <- bitr(names(ranked), fromType = "ENSEMBL", toType = "ENTREZID",
            OrgDb = org.Gg.eg.db)
ids <- ids[!duplicated(ids$ENTREZID), ]

rk2  <- ranked[ids$ENSEMBL]
names(rk2) <- ids$ENTREZID
rk2  <- sort(rk2, decreasing = TRUE)

kk2 <- gseKEGG(geneList     = rk2,
               organism     = "gga",
               minGSSize    = 3,
               maxGSSize    = 800,
               pvalueCutoff = 0.05,
               pAdjustMethod = "BH",
               keyType       = "ncbi-geneid")

pdf("KEGG_GSEA_dotplot.pdf")
print(dotplot(kk2, showCategory = 10, split = ".sign") +
        facet_grid(. ~ .sign))
dev.off()
#```

## 11  Pathway view examples (optional – KEGG API)
#```{r pathview, eval = FALSE}
# pathview(gene.data = rk2, pathway.id = "gga04260", species = "gga")
#```

## 12  Session info (for reproducibility)
```{r session-info}
writeLines(capture.output(sessionInfo()), "sessionInfo.txt")
```
