---
title: "DESeq2_Clemson_output GOS refactor – UPDATED (2025-06-30)"
author: "Rooksie Noorai (original) → Galen O'Shea-Stone (refactor 6/30)"
date: "2025-06-30"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(666)
output_dir <- "deseq2_outputs"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
```

## 1  Load libraries
```{r libs}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

library(DESeq2)
library(apeglm)
library(clusterProfiler)
library(enrichplot)
library(AnnotationDbi)
library(org.Gg.eg.db)
library(pathview)
library(ggplot2)
```

## 2  Read counts & metadata
```{r read-counts}
Expression <- read.delim("redo.ensembl.stranded.counts.renamed.subset.txt")
rownames(Expression) <- Expression$Geneid
cts <- as.matrix(Expression[ , -1]); storage.mode(cts) <- "integer"
condition <- factor(c("controlF","controlF","experimental","controlP",
                      "experimental","experimental","controlP",
                      "controlF","controlP","controlP"),
                    levels = c("controlF", "controlP", "experimental"))
coldata <- data.frame(condition = condition, row.names = colnames(cts))
```

## 3  DESeqDataSet & pre-filter
```{r dds-setup}
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData   = coldata,
                              design    = ~ condition)
keep <- rowSums(counts(dds) >= 10) >= 3
dds  <- dds[keep, ]
```

## 4  Run DESeq2
```{r deseq-run}
dds <- DESeq(dds)
resultsNames(dds)
```

## 5  QC plots (VST)
```{r vst-qc, fig.width = 6.5, fig.height = 5}
vsd <- vst(dds, blind = FALSE)

pca_plot <- plotPCA(vsd, intgroup = "condition") +
            ggtitle("PCA – variance-stabilised")
print(pca_plot)

ggsave(file.path(output_dir, "DESeq2_PCA_plot.pdf"), pca_plot, width = 6, height = 5)

pdf(file.path(output_dir, "size_factors_barplot.pdf"))
barplot(sizeFactors(dds), main = "Normalisation size factors", las = 2, ylab = "sizeFactor")
dev.off()

pdf(file.path(output_dir, "dispersion_trend.pdf"))
plotDispEsts(dds)
dev.off()
```

## 6  Experimental vs controlF
```{r primary-contrast}
res_raw <- results(dds, name = "condition_experimental_vs_controlF")
res_shr <- lfcShrink(dds, coef = "condition_experimental_vs_controlF", type = "apeglm")
res_final <- data.frame(baseMean = res_raw$baseMean,
                        log2FC_shrunken = res_shr$log2FoldChange,
                        lfcSE_shrunken = res_shr$lfcSE,
                        pvalue = res_raw$pvalue,
                        padj   = res_raw$padj,
                        row.names = rownames(res_raw))
write.csv(res_final, file.path(output_dir, "DESeq2_experimental_vs_controlF.csv"))
```

## 7  Experimental vs controlP
```{r secondary-contrast}
dds_P <- dds; dds_P$condition <- relevel(dds_P$condition, ref = "controlP"); dds_P <- DESeq(dds_P)
res_P_shr <- lfcShrink(dds_P, coef = "condition_experimental_vs_controlP", type = "apeglm")
core_cols_P <- intersect(c("baseMean","log2FoldChange","lfcSE","stat","pvalue","padj","svalue"), colnames(res_P_shr))
write.csv(as.data.frame(res_P_shr)[ , core_cols_P], file.path(output_dir, "DESeq2_experimental_vs_controlP.csv"))
```

## 8  ControlP vs controlF
```{r controlP-vs-controlF}
res_CF_raw <- results(dds, name = "condition_controlP_vs_controlF")
res_CF_shr <- lfcShrink(dds, coef = "condition_controlP_vs_controlF", type = "apeglm")
res_CF <- data.frame(baseMean = res_CF_raw$baseMean,
                     log2FC_shrunken = res_CF_shr$log2FoldChange,
                     lfcSE_shrunken  = res_CF_shr$lfcSE,
                     pvalue = res_CF_raw$pvalue,
                     padj   = res_CF_raw$padj,
                     row.names = rownames(res_CF_raw))
write.csv(res_CF, file.path(output_dir, "DESeq2_controlP_vs_controlF.csv"))
```

## 9  Export normalised counts for reviewers
```{r export-counts}
write.csv(counts(dds, normalized = TRUE), file.path(output_dir, "normalised_counts.csv"))
```

---
###  GSEA blocks (commented; edit & set `eval=TRUE` to run)
```{r gsea-go, eval = FALSE}
ranked <- res_shr$log2FoldChange; names(ranked) <- rownames(res_shr)
ranked <- sort(na.omit(ranked), decreasing = TRUE)

# GSEA (BH adjusted)
gse_go <- gseGO(geneList     = ranked,
                keyType      = "ENSEMBL",
                ont          = "ALL",
                OrgDb        = org.Gg.eg.db,
                nPerm        = 10000,
                minGSSize    = 3,
                maxGSSize    = 800,
                pvalueCutoff = 0.05,
                pAdjustMethod = "BH",
                verbose      = TRUE)

pdf("GO_GSEA_dotplot.pdf")
print(dotplot(gse_go, showCategory = 10, split = ".sign") +
        facet_grid(. ~ .sign))
dev.off()
#```

## 10  Gene-set enrichment – KEGG
#```{r gsea-kegg}
# Convert ENSEMBL → ENTREZ
ids <- bitr(names(ranked), fromType = "ENSEMBL", toType = "ENTREZID",
            OrgDb = org.Gg.eg.db)
ids <- ids[!duplicated(ids$ENTREZID), ]

rk2  <- ranked[ids$ENSEMBL]
names(rk2) <- ids$ENTREZID
rk2  <- sort(rk2, decreasing = TRUE)

kk2 <- gseKEGG(geneList     = rk2,
               organism     = "gga",
               minGSSize    = 3,
               maxGSSize    = 800,
               pvalueCutoff = 0.05,
               pAdjustMethod = "BH",
               keyType       = "ncbi-geneid")

pdf("KEGG_GSEA_dotplot.pdf")
print(dotplot(kk2, showCategory = 10, split = ".sign") +
        facet_grid(. ~ .sign))
dev.off()
#```

## 11  Pathway view examples (optional – KEGG API)     uncomment if wanted (enrichment_analysis script takes over in this pipeline)
#```{r pathview, eval = FALSE}
# pathview(gene.data = rk2, pathway.id = "gga04260", species = "gga")
#```
```


## 12  Session info (for reproducibility)
```{r session-info}
writeLines(capture.output(sessionInfo()), "sessionInfo.txt")
```
