---
title: "cGAS-STING Pathway Map Notebook (v6)"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1) Set Parameters

Edit these values if needed, then click **Run All** (or **Knit**).

```{r parameters}
# Input DESeq2 CSV (Ensembl IDs in first column/ensembl_id).
input_file <- "DESeq2_GRCg7b_Sacral_vs_Free_by_ENS.csv"

# Output directory for map + tables.
output_dir <- "cgast_sting_out"

# Plot title.
plot_title <- "Sacral vs Free (cGAS-STING map)"

# Run apoptosis/necroptosis/combined maps too.
run_extra_maps <- TRUE
```

## 2) Load Script

```{r load-script, echo=FALSE}
# Find and source the script in the same folder as this notebook.
script_candidates <- c(
  "cgast_sting_map.R",
  file.path(getwd(), "cgast_sting_map.R"),
  "cgast_sting_map_v2.R",
  file.path(getwd(), "cgast_sting_map_v2.R")
)
script_path <- script_candidates[file.exists(script_candidates)][1]

if (is.na(script_path) || !nzchar(script_path)) {
  stop("Could not find cgast_sting_map.R in the notebook working directory.")
}

source(script_path)
```

## 3) Run Pathway Map

```{r run-map, echo=FALSE}
if (!file.exists(input_file)) {
  stop("Input file not found: ", input_file)
}

if (!exists("run_cgast_sting_map")) {
  stop("run_cgast_sting_map() not found after sourcing cgast_sting_map.R")
}

results <- run_cgast_sting_map(
  input = input_file,
  outdir = output_dir,
  title = plot_title
)

results
```

## 4) Preview Outputs

This uses the current script settings (white background, black arrows, text-fitted boxes, capped arrows, and updated significance legend).

```{r preview-image, echo=FALSE, out.width="100%"}
knitr::include_graphics(results$png)
```

```{r preview-tables, echo=FALSE}
nodes_preview <- readr::read_tsv(results$nodes, show_col_types = FALSE)
missing_preview <- readr::read_tsv(results$missing, show_col_types = FALSE)
all_results_preview <- readr::read_csv(results$all_results, show_col_types = FALSE)

head(nodes_preview, 10)
head(missing_preview, 10)
head(all_results_preview, 10)
```

## 5) Export Results CSV (Notebook)

This writes a notebook-generated CSV copy of the full annotated results table.

```{r export-results-csv, echo=FALSE}
notebook_results_csv <- file.path(output_dir, "cgast_sting_results_from_notebook.csv")
readr::write_csv(all_results_preview, notebook_results_csv, na = "NA")
notebook_results_csv
```

## 6) Output Files

```{r output-paths, echo=FALSE}
unlist(c(results, list(notebook_results_csv = notebook_results_csv)))
```

## 7) Session Info

```{r session-info, echo=TRUE}
session_info_file <- file.path(output_dir, "sessionInfo.txt")
session_text <- capture.output(sessionInfo())
writeLines(session_text, session_info_file, useBytes = TRUE)
cat(paste(session_text, collapse = "\n"))
cat("\n\nSaved session info to: ", session_info_file, "\n", sep = "")
```

## 8) Additional Pathway Maps

This chunk runs by default when `run_extra_maps <- TRUE`.
This version also writes an extra additive file:
`combined_death_multipanel.png` (Panel A = pathway map, Panel B = interpretation/validation notes).

```{r extra-pathways, echo=TRUE}
if (isTRUE(run_extra_maps)) {
  extra <- run_pathway_maps(
    input = input_file,
    outdir = output_dir,
    title = plot_title,
    pathways = c("apoptosis", "necroptosis"),
    include_combined = TRUE
  )
  extra
} else {
  extra <- NULL
  cat("Skipping extra pathway maps because run_extra_maps is FALSE.\n")
}
```

```{r preview-extra-pathways, echo=FALSE, out.width="100%"}
if (!is.null(extra)) {
  knitr::include_graphics(c(
    extra$apoptosis$png,
    extra$necroptosis$png,
    extra$combined_death$png
  ))
}
```

```{r preview-combined-multipanel, echo=FALSE, out.width="100%"}
if (!is.null(extra)) {
  knitr::include_graphics(extra$combined_death$multipanel_png)
}
```
