---
title: "Pathway Mapper v3 — Control Notebook"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Overview

This notebook controls `pathway_mapper_v3.R`. It sources the script, lets
you configure inputs and options, and runs pathway maps individually or in
batch. All output files land in the directory you set in **Section 2**.

---

# 1. Source the script

Run this chunk first every session. It loads all functions and the pathway
registry into your environment.

```{r source}
source("pathway_mapper_v4.R")
```

---

# 2. Configure inputs

Edit these three variables. Everything else in the notebook reads from them.

```{r config}
# ── Required ──────────────────────────────────────────────────────────────
INPUT_CSV <- "DESeq2_Free_vs_Pygostyle_by_ENS.csv"  # path to DESeq2 CSV
OUTDIR    <- "pathway_out"                                # output directory
TITLE     <- "Sacral vs Free"                             # plot title prefix

# ── Optional ──────────────────────────────────────────────────────────────
P_CUTOFF  <- 0.05                                   # significance threshold
LFC_COLS  <- c("log2FC_shrunken", "log2FC")         # fold-change column priority
ID_COLS   <- c("ensembl_id", "Unnamed: 0", "gene", "id")  # Ensembl ID column candidates
```

> **Tip:** Use `file.choose()` in the console to pick a file interactively,
> then paste the path above.

---

# 3. List available pathways

```{r list-pathways}
list_pathways()
```

---

# 4. Run a single pathway

Change `PATHWAY` to any name from `list_pathways()`.
Available pathways now include: `"oxysterol"` (Oxysterol Signaling in Bone Fusion).

```{r single-pathway}
PATHWAY <- "cgas_sting"   # ← change this  (e.g. "oxysterol")

run_pathway_maps(
  input    = INPUT_CSV,
  outdir   = OUTDIR,
  title    = TITLE,
  pathways = PATHWAY,
  p_cutoff = P_CUTOFF,
  lfc_column_preference = LFC_COLS,
  id_column_candidates  = ID_COLS,
  stitch_panels = FALSE   # no overview grid when running just one pathway
)
```

---

# 5. Run a custom selection of pathways

```{r multi-pathway}
MY_PATHWAYS <- c("apoptosis", "necroptosis", "oxysterol")

run_pathway_maps(
  input    = INPUT_CSV,
  outdir   = OUTDIR,
  title    = TITLE,
  pathways = MY_PATHWAYS,
  p_cutoff = P_CUTOFF,
  lfc_column_preference = LFC_COLS,
  id_column_candidates  = ID_COLS,
  stitch_panels = TRUE   # saves all_pathways_overview.png
)
```

---

# 6. Run all pathways

Renders every pathway in the registry and stitches them into
`all_pathways_overview.png`.

```{r all-pathways}
run_pathway_maps(
  input    = INPUT_CSV,
  outdir   = OUTDIR,
  title    = TITLE,
  pathways = list_pathways(),
  p_cutoff = P_CUTOFF,
  lfc_column_preference = LFC_COLS,
  id_column_candidates  = ID_COLS,
  stitch_panels = TRUE
)
```

---

# 7. Preview outputs inline

After running any section above, use these chunks to view the results
directly in the notebook.

## Individual pathway map

```{r preview-map, fig.width=14, fig.height=7.5}
PREVIEW_PATHWAY <- "cgas_sting"   # ← change to whichever pathway you want

cfg         <- get_pathway_config(PREVIEW_PATHWAY)
prepared_de <- prepare_de_table_for_pathways(
  input                 = INPUT_CSV,
  lfc_column_preference = LFC_COLS,
  id_column_candidates  = ID_COLS
)

result <- render_pathway_map(
  config                       = cfg,
  prepared_de                  = prepared_de,
  outdir                       = OUTDIR,
  title                        = TITLE,
  p_cutoff                     = P_CUTOFF,
  show_compartments            = TRUE,
  min_nodes_per_compartment    = 2,
  force_singleton_compartments = c("Plasma membrane", "ER/Golgi"),
  return_plot                  = TRUE
)

result$plot
```

## Interpretation panel

```{r preview-panel, fig.width=6, fig.height=7.5}
make_interpretation_panel(
  nodes_annot      = result$node_tbl,
  edges_cfg        = cfg$edges,
  title            = paste0(cfg$display_name, " \u2014 Interpretation"),
  validation_notes = cfg$validation_notes
)
```

## Node table

```{r preview-nodes}
result$node_tbl |>
  dplyr::select(node_id, label, module, compartment,
                log2FC_used, padj, node_status) |>
  dplyr::arrange(node_status, dplyr::desc(abs(log2FC_used))) |>
  knitr::kable(digits = 3)
```

## Missing genes

```{r preview-missing}
missing <- result$node_tbl |>
  dplyr::filter(node_status == "Missing gene") |>
  dplyr::select(node_id, label, module, compartment)

if (nrow(missing) == 0) {
  cat("No missing genes — all pathway nodes matched in dataset.\n")
} else {
  knitr::kable(missing)
}
```

---

# 8. Inspect a pathway config

Useful for checking node positions, edge wiring, or synonyms before
running a map.

```{r inspect-config}
INSPECT <- "inflammasome"   # ← change to any pathway name

cfg <- get_pathway_config(INSPECT)

cat("=== Pathway:", cfg$display_name, "===\n\n")
cat("Nodes (", nrow(cfg$nodes), "):\n")
print(cfg$nodes)

cat("\nEdges (", nrow(cfg$edges), "):\n")
print(cfg$edges)

cat("\nSynonyms:\n")
print(cfg$synonyms)

cat("\nValidation notes:\n")
cat(paste0("  - ", cfg$validation_notes, collapse = "\n"), "\n")
```

---

# 9. List output files

```{r list-outputs}
if (dir.exists(OUTDIR)) {
  files <- list.files(OUTDIR, full.names = FALSE)
  data.frame(file = sort(files)) |> knitr::kable()
} else {
  cat("Output directory does not exist yet — run a pathway first.\n")
}
```

---

# Session info

```{r session-info}
sessionInfo()
```
