---
title: "DESeq2_Clemson_output – full QC + stats (2025‑06‑30)"
author: "Rooksie Noorai (orig) → Galen O'Shea‑Stone (refactor + QC)"
date: "2025‑06‑30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(666)
output_dir <- "deseq2_outputs"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
```

## 1  Load libraries
```{r libs}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# Core analysis packages ------------------------------------------------
core_pkgs <- c("DESeq2", "apeglm", "AnnotationDbi", "org.Gg.eg.db")
for (pkg in core_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE))
    BiocManager::install(pkg, ask = FALSE)
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}

# QC / visualisation packages ------------------------------------------
qc_pkgs <- c("ggplot2", "pheatmap", "vsn")
for (pkg in qc_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE))
    BiocManager::install(pkg, ask = FALSE)
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}

if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# core analysis libs (install first run only)
# BiocManager::install(c("DESeq2","apeglm","clusterProfiler","enrichplot",
#                        "org.Gg.eg.db","AnnotationDbi","pathview",
#                        "pheatmap","vsn"), ask = FALSE)

library(DESeq2)
library(apeglm)
library(AnnotationDbi)
library(org.Gg.eg.db)

# visualisation / QC helper libraries
Libraries <- c("ggplot2", "pheatmap", "vsn")
for (pkg in Libraries)
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
```

## 2  Read counts & metadata
```{r read-counts}
Expression <- read.delim("redo.ensembl.stranded.counts.renamed.subset.txt")
rownames(Expression) <- Expression$Geneid
cts <- as.matrix(Expression[ , -1]); storage.mode(cts) <- "integer"
condition <- factor(c("controlF","controlF","experimental","controlP",
                      "experimental","experimental","controlF",
                      "controlP","controlP","controlP"),
                    levels = c("controlF","controlP","experimental"))
coldata <- data.frame(condition = condition, row.names = colnames(cts))
```

## 3  DESeqDataSet & pre‑filter
```{r dds-setup}
dds <- DESeqDataSetFromMatrix(countData = cts, colData = coldata, design = ~ condition)
keep <- rowSums(counts(dds) >= 10) >= 3   # ≥10 reads in ≥3 samples
dds  <- dds[keep, ]
```

## 4  Run DESeq2
```{r deseq-run}
dds <- DESeq(dds)
resultsNames(dds)
```

## 5  Variance‑stabilised QC
```{r vst-qc, fig.width=6.5, fig.height=5}
vsd <- vst(dds, blind = FALSE)

# PCA
pca_plot <- plotPCA(vsd, intgroup = "condition") + ggtitle("PCA – VST counts")
print(pca_plot)

ggsave(file.path(output_dir, "PCA_vst.pdf"), pca_plot, width = 6, height = 5)

# Size factors
pdf(file.path(output_dir, "size_factors.pdf"))
barplot(sizeFactors(dds), las = 2, ylab = "sizeFactor", main = "Size factors")
dev.off()

# Dispersion trend
pdf(file.path(output_dir, "dispersion_trend.pdf"))
plotDispEsts(dds)
dev.off()
```

## 6  Additional validation metrics
```{r qc-extra}
# 6.1  Sample‑distance heatmap
mat <- as.matrix(dist(t(assay(vsd))))
pdf(file.path(output_dir, "sample_distance_heatmap.pdf"))
pheatmap(mat, clustering_distance_rows = "euclidean", clustering_distance_cols = "euclidean",
         main = "Sample–sample distance (VST)")
dev.off()

# 6.2  Cook's distance boxplot
pdf(file.path(output_dir, "cooks_distance_boxplot.pdf"))
boxplot(log10(assays(dds)[["cooks"]]), range = 0, outline = FALSE,
        las = 2, ylab = "log10 Cook's distance", main = "Influential observations")
dev.off()

# 6.3  Mean–SD plot of VST counts
pdf(file.path(output_dir, "meanSD_vst.pdf"))
vsn::meanSdPlot(assay(vsd))
dev.off()

# 6.4  P‑value histogram
res_tmp <- results(dds)
pdf(file.path(output_dir, "pvalue_histogram.pdf"))
hist(res_tmp$pvalue, breaks = 50, col = "grey", main = "P-value distribution", xlab = "p-value")
dev.off()

# 6.5  Replicate correlation pairs plot (quick visual)
pdf(file.path(output_dir, "replicate_correlation_pairs.pdf"))
pairs(log10(counts(dds, norm = TRUE) + 1), pch = ".", main = "Normalised count pairs")
dev.off()
```

## 7  Experimental vs controlF
```{r exp-vs-ctrlF}
res_raw <- results(dds, name = "condition_experimental_vs_controlF")
res_shr <- lfcShrink(dds, coef = "condition_experimental_vs_controlF", type = "apeglm")

# MA plots (raw & shrunk)
pdf(file.path(output_dir, "MA_raw.pdf")); plotMA(res_raw, alpha = 0.05); dev.off()
pdf(file.path(output_dir, "MA_shrunk.pdf")); plotMA(res_shr, alpha = 0.05); dev.off()

res_final <- data.frame(baseMean        = res_raw$baseMean,
                        log2FC_shrunken = res_shr$log2FoldChange,
                        lfcSE_shrunken  = res_shr$lfcSE,
                        pvalue          = res_raw$pvalue,
                        padj            = res_raw$padj,
                        row.names       = rownames(res_raw))
write.csv(res_final, file.path(output_dir, "DESeq2_experimental_vs_controlF.csv"))

# Up/down counts summary
sig <- res_final$padj < 0.05 & !is.na(res_final$padj)
up   <- sum(sig & res_final$log2FC_shrunken > 0)
down <- sum(sig & res_final$log2FC_shrunken < 0)
write.table(data.frame(up = up, down = down),
            file = file.path(output_dir, "sig_gene_counts_experimental_vs_controlF.txt"),
            sep = "\t", row.names = FALSE)
```

## 8  Experimental vs controlP
```{r exp-vs-ctrlP}
dds_P <- dds; dds_P$condition <- relevel(dds_P$condition, ref = "controlP"); dds_P <- DESeq(dds_P)
res_P_shr <- lfcShrink(dds_P, coef = "condition_experimental_vs_controlP", type = "apeglm")
core_cols_P <- intersect(c("baseMean","log2FoldChange","lfcSE","stat","pvalue","padj","svalue"), colnames(res_P_shr))
write.csv(as.data.frame(res_P_shr)[ , core_cols_P], file.path(output_dir, "DESeq2_experimental_vs_controlP.csv"))
```

## 9  ControlF vs controlP
```{r ctrlP-vs-ctrlF}
# make a new DDS where "controlP" is the reference
dds_CF <- dds
dds_CF$condition <- relevel(dds_CF$condition, ref = "controlP")
dds_CF <- DESeq(dds_CF)

# now resultsNames(dds_CF) will include "condition_controlF_vs_controlP"
resultsNames(dds_CF)
# [1] "condition_controlP_vs_experimental"
# [2] "condition_controlF_vs_controlP"
# [3] "condition_experimental_vs_controlP"

# use the coef name directly with lfcShrink(type="apeglm")
res_CF_raw <- results(dds_CF, name = "condition_controlF_vs_controlP")
res_CF_shr <- lfcShrink(dds_CF,
                        coef = "condition_controlF_vs_controlP",
                        type = "apeglm")

# assemble & write out as before
res_CF <- data.frame(baseMean        = res_CF_raw$baseMean,
                     log2FC_shrunken = res_CF_shr$log2FoldChange,
                     lfcSE_shrunken  = res_CF_shr$lfcSE,
                     pvalue          = res_CF_raw$pvalue,
                     padj            = res_CF_raw$padj,
                     row.names       = rownames(res_CF_raw))
write.csv(res_CF, file.path(output_dir, "DESeq2_controlF_vs_controlP.csv"))

```

## 10  Export normalised counts
```{r export-counts}
write.csv(counts(dds, normalized = TRUE), file.path(output_dir, "normalised_counts.csv"))
```

---
###  GSEA blocks (commented; edit & set `eval=TRUE` to run)
```{r gsea-go, eval = FALSE}
ranked <- res_shr$log2FoldChange; names(ranked) <- rownames(res_shr)
ranked <- sort(na.omit(ranked), decreasing = TRUE)

# GSEA (BH adjusted)
gse_go <- gseGO(geneList     = ranked,
                keyType      = "ENSEMBL",
                ont          = "ALL",
                OrgDb        = org.Gg.eg.db,
                nPerm        = 10000,
                minGSSize    = 3,
                maxGSSize    = 800,
                pvalueCutoff = 0.05,
                pAdjustMethod = "BH",
                verbose      = TRUE)

pdf("GO_GSEA_dotplot.pdf")
print(dotplot(gse_go, showCategory = 10, split = ".sign") +
        facet_grid(. ~ .sign))
dev.off()
#```

## 10  Gene-set enrichment – KEGG
#```{r gsea-kegg}
# Convert ENSEMBL → ENTREZ
ids <- bitr(names(ranked), fromType = "ENSEMBL", toType = "ENTREZID",
            OrgDb = org.Gg.eg.db)
ids <- ids[!duplicated(ids$ENTREZID), ]

rk2  <- ranked[ids$ENSEMBL]
names(rk2) <- ids$ENTREZID
rk2  <- sort(rk2, decreasing = TRUE)

kk2 <- gseKEGG(geneList     = rk2,
               organism     = "gga",
               minGSSize    = 3,
               maxGSSize    = 800,
               pvalueCutoff = 0.05,
               pAdjustMethod = "BH",
               keyType       = "ncbi-geneid")

pdf("KEGG_GSEA_dotplot.pdf")
print(dotplot(kk2, showCategory = 10, split = ".sign") +
        facet_grid(. ~ .sign))
dev.off()
#```

## 11  Pathway view examples (optional – KEGG API)     uncomment if wanted (enrichment_analysis script takes over in this pipeline)
#```{r pathview, eval = FALSE}
# pathview(gene.data = rk2, pathway.id = "gga04260", species = "gga")
#```
```

## 11  Session info
```{r session-info}
writeLines(capture.output(sessionInfo()), file.path(output_dir, "sessionInfo.txt"))
```

