---
title: "Pygo_gene_enrichment (AUTO-WRAP, robust – KEGGREST)"
author: "Galen O'Shea-Stone"
date: "2025-06-11"
output: html_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

##############################################################################
## Transcriptomics → GO / KEGG (via KEGGREST) → Reactome enrichment (Gallus) ##
## + diagnostics, Excel export, auto-wrapped dot-plots (robust)             ##
##############################################################################

## 0. CONFIGURATION -----------------------------------------------------------
input_file   <- "experimental_vs_controlF_forR.csv"   # DE table: gene, logFC, padj
species_name <- "chicken"         # ReactomePA shorthand
kegg_code    <- "gga"             # KEGG organism code
padj_cutoff  <- 0.05              # adjusted-P significance threshold
logFC_cutoff <- 2                 # minimum |log2FC| to keep
topN         <- 2000              # keep top N up + N down genes

## 1. INSTALL & LOAD PACKAGES -------------------------------------------------
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

cran_pkgs <- c("UpSetR", "EnhancedVolcano", "stringr", "glue")
bio_pkgs  <- c("clusterProfiler", "ReactomePA", "pathview", "biomaRt",
               "AnnotationDbi", "dplyr", "ggplot2", "openxlsx",
               "ComplexHeatmap", "circlize", "enrichplot", "KEGGREST")

for (p in cran_pkgs)
  if (!requireNamespace(p, quietly = TRUE))
    install.packages(p, dependencies = TRUE)

for (p in bio_pkgs)
  if (!requireNamespace(p, quietly = TRUE))
    BiocManager::install(p, ask = FALSE, update = FALSE)

if (!requireNamespace("org.Gg.eg.db", quietly = TRUE))
    BiocManager::install("org.Gg.eg.db", ask = FALSE, update = FALSE)

library(clusterProfiler); library(ReactomePA);  library(pathview)
library(biomaRt);          library(AnnotationDbi); library(dplyr)
library(ggplot2);          library(openxlsx);       library(ComplexHeatmap)
library(circlize);         library(UpSetR);         library(EnhancedVolcano)
library(enrichplot);       library(stringr);        library(glue)
library(KEGGREST);         library(org.Gg.eg.db);    library(tidyr)

orgDb <- org.Gg.eg.db

## ---- Neural keyword helpers (define once, reuse everywhere) ---------------
neural_kw       <- c("neuro","neuron","neura","neural","synap","axon","dendrit","neur","brain","nerv",
                     "glia","myelin","schwann","spine","plasticity","cerebral",
                     "cerebell","cortex","spinal","ganglia","astrocyte",
                     "oligodendro","hippocamp","amygdala")

neural_class_kw <- c("Nervous system","Neurodegenerative","Sensory system")

neural_pattern  <- paste(c(neural_kw, neural_class_kw), collapse = "|")

neural_kw_extra <- c(
  # signalling
  "neurotransm", "neuropept", "neuromodul", 
  # peripheral/autonomic
  "autonom", "sympathetic", "parasympathetic", "neuromuscular",
  # development
  "neurogenesis", "axonogen", "myelination", 
  # glial
  "microglia"
)
neural_kw <- unique(c(neural_kw, neural_kw_extra))
neural_pattern <- paste(c(neural_kw, neural_class_kw), collapse = "|")



dir.create("output", showWarnings = FALSE)
## 2. READ & PREP DATA  -------------------------------------------------------
tbl_raw <- read.csv(input_file, check.names = FALSE)
tbl     <- tbl_raw |>
  rename(gene = 1, logFC = 2, padj = 3)

## --- New: robust ID-type detection & cleaning -------------------------------
# Strip Ensembl version numbers such as ‘.1’, ‘.2’ if present
tbl$gene <- sub("\\.\\d+$", "", tbl$gene)

# Decide whether we are holding Ensembl IDs or gene symbols
is_ensembl <- all(grepl("^ENS", tbl$gene, ignore.case = FALSE))
keytype    <- if (is_ensembl) "ENSEMBL" else "SYMBOL"

message("Detected keytype → ", keytype)

## --- Map to ENTREZ (and get SYMBOLs for plotting) ---------------------------
# For Ensembl input we request both ENTREZ and SYMBOL; for symbol input we also
# keep the original symbol and recover the Ensembl ID if you ever need it.
map_cols <- if (keytype == "ENSEMBL") c("ENTREZID", "SYMBOL") else c("ENTREZID", "ENSEMBL")

gene_map <- AnnotationDbi::select(
  orgDb,
  keys     = tbl$gene,
  columns  = map_cols,
  keytype  = keytype
)

# Merge mapping back, keeping the original input column name ‘gene’
tbl <- left_join(tbl, gene_map,
                 by = setNames(keytype, "gene")) |>
       filter(!is.na(ENTREZID))

## --- A tidy ‘gene_label’ helper ---------------------------------------------
# If SYMBOL wasn’t returned (e.g. because the input genes are already symbols),
# create an empty placeholder so coalesce() never errors.
if (!"SYMBOL" %in% names(tbl))
    tbl$SYMBOL <- NA_character_

tbl <- tbl %>%
    mutate(gene_label = dplyr::coalesce(SYMBOL, gene))

## 2a. Define universe & test set --------------------------------------------
gene_universe <- unique(as.character(tbl$ENTREZID))

tbl_sig <- tbl |>
  filter(padj < padj_cutoff, abs(logFC) >= logFC_cutoff) |>
  arrange(desc(logFC))
if (nrow(tbl_sig) > 2 * topN)
    tbl_sig <- bind_rows(head(tbl_sig, topN), tail(tbl_sig, topN))

sig_genes <- unique(as.character(tbl_sig$ENTREZID))

## 3. SUMMARY STATS -----------------------------------------------------------
num_input    <- nrow(tbl_raw)
num_mapped   <- nrow(tbl)
num_unmapped <- num_input - num_mapped
num_universe <- length(gene_universe)
num_sig      <- length(sig_genes)
pct_sig      <- 100 * num_sig / num_universe

cat(sprintf(
  "\n====== SUMMARY ======\nInput genes:    %d\nMapped genes:   %d\nUnmapped genes: %d\nUniverse:       %d\nTest set:       %d (%.1f%% of universe)\n",
  num_input, num_mapped, num_unmapped, num_universe, num_sig, pct_sig))

summary_stats <- data.frame(
  Step       = c("Input genes", "Mapped genes", "Unmapped genes",
                 "Universe", "Test set"),
  Gene_Count = c(num_input, num_mapped, num_unmapped, num_universe, num_sig),
  Percent    = c(NA,
                 100 * num_mapped / num_input,
                 100 * num_unmapped / num_input,
                 NA,
                 pct_sig)
)
write.xlsx(summary_stats,
           file = "output/summary_stats.xlsx", overwrite = TRUE)

## Pie chart ------------------------------------------------------------------
pdf("output/testset_vs_universe_pie.pdf", 5, 5)
pie(c(num_sig, num_universe - num_sig),
    labels = c(sprintf("Test set\n(%d)", num_sig),
               sprintf("Universe\n(%d)", num_universe - num_sig)),
    col = c("deepskyblue", "grey90"),
    main = "Genes in Test Set vs. Universe")
dev.off()

## Barplot --------------------------------------------------------------------
bar_vals <- c(Input = num_input, Mapped = num_mapped,
              Universe = num_universe, TestSet = num_sig)
pdf("output/gene_counts_barplot.pdf", 5, 4)
barplot(bar_vals,
        main = "Gene Counts at Key Steps",
        ylab = "Number of genes",
        col  = "darkseagreen",
        ylim = c(0, max(bar_vals) * 1.1))
dev.off()

## ---- 3b  Up- vs down-regulated summary -------------------------------------
## Uses the same padj_cutoff & logFC_cutoff you set earlier

tbl <- tbl %>%                        # tbl already has gene, logFC, padj
  mutate(Regulation = dplyr::case_when(
    logFC >=  logFC_cutoff  & padj < padj_cutoff ~ "Up",
    logFC <= -logFC_cutoff  & padj < padj_cutoff ~ "Down",
    TRUE                                         ~ "NoChange"
  ))

# Counts & percentages
up_down_counts <- tbl %>%
  filter(Regulation %in% c("Up", "Down")) %>%
  count(Regulation, name = "Gene_Count") %>%
  mutate(Percent = round(100 * Gene_Count / sum(Gene_Count), 1))

cat("\n======  UP vs DOWN (significant)  ======\n")
print(up_down_counts)

# Full tables for convenient inspection
up_sig   <- tbl %>% filter(Regulation == "Up")   %>% arrange(desc(logFC))
down_sig <- tbl %>% filter(Regulation == "Down") %>% arrange(logFC)

openxlsx::write.xlsx(list(
  UpDown_Counts = up_down_counts,
  Up_Genes      = up_sig,
  Down_Genes    = down_sig
),
file = "output/up_vs_down_summary.xlsx", overwrite = TRUE)

message("✓ up_vs_down_summary.xlsx written to output/")

## ---- helpers_plotting, echo=FALSE, message=FALSE ---------------------------
save_plot <- function(p, fname, w = 7, h = 5)
  ggsave(file.path("output", fname), plot = p,
         width = w, height = h, units = "in")

make_dot <- function(res, stub, title,
                     n_cat = 15, wrap_width = 45,
                     text_size = 9, text_angle = 0,
                     w = 7, h = 5) {
  df <- as.data.frame(res)
  if (nrow(df) == 0) return(invisible())
  df <- df[seq_len(min(n_cat, nrow(df))), ]
  df$Description <- stringr::str_wrap(df$Description, width = wrap_width)

  if (all(c("GeneRatio", "Count") %in% names(df))) {
    df$GeneRatioNum <- vapply(strsplit(df$GeneRatio, "/"),
                              function(x) as.numeric(x[1]) / as.numeric(x[2]),
                              FUN.VALUE = numeric(1))
    x_var <- "GeneRatioNum"; x_lab <- "Gene ratio"; size_var <- "Count"
  } else if (all(c("NES", "setSize") %in% names(df))) {
    x_var <- "NES";          x_lab <- "NES";        size_var <- "setSize"
  } else stop("Unsupported enrichment object")

  p <- ggplot(df, aes_string(x = x_var,
                             y = paste0("reorder(Description,", x_var, ")"),
                             size = size_var,
                             colour = "p.adjust")) +
    geom_point() +
    scale_colour_continuous(low = "red", high = "blue", name = "adj-P") +
    scale_size(range = c(2, 6)) +
    labs(title = title, x = x_lab, y = NULL) +
    theme_bw() +
    theme(axis.text.y = element_text(size  = text_size,
                                     angle = text_angle,
                                     hjust = ifelse(text_angle == 0, 1, 0.5)))
  save_plot(p, sprintf("dotplot_%s.pdf", stub), w, h)
}


## 4. ENRICHMENT ANALYSES -----------------------------------------------------
### 4.1 GO BP ORA
ego <- enrichGO(
  gene          = sig_genes,
  OrgDb         = orgDb,
  keyType       = "ENTREZID",
  ont           = "BP",
  universe      = gene_universe,
  pAdjustMethod = "BH",
  readable      = TRUE
)

## 4.1b GO MF ORA  -----------------------------------------------------------
ego_MF <- enrichGO(
  gene          = sig_genes,
  OrgDb         = orgDb,
  keyType       = "ENTREZID",
  ont           = "MF",
  universe      = gene_universe,
  pAdjustMethod = "BH",
  readable      = TRUE
)

### 4.2 KEGG ORA (via KEGGREST + enricher) -----------------------------------
kmap <- KEGGREST::keggLink("pathway", kegg_code)   # gene → pathway
kegg_term2gene <- data.frame(
  term = sub("path:",  "", unname(kmap)),
  gene = sub(paste0(kegg_code, ":"), "", names(kmap)),
  stringsAsFactors = FALSE
)

kdesc <- KEGGREST::keggList("pathway", kegg_code)
kegg_term2name <- data.frame(
  term = sub("path:", "", names(kdesc)),
  name = sub(" - .*", "", unname(kdesc)),
  stringsAsFactors = FALSE
)

ekegg <- enricher(
  gene          = sig_genes,
  universe      = gene_universe,
  TERM2GENE     = kegg_term2gene,
  TERM2NAME     = kegg_term2name,
  pAdjustMethod = "BH"
)
if (nrow(as.data.frame(ekegg)) > 0)
    ekegg <- setReadable(ekegg, orgDb, keyType = "ENTREZID")

### 4.3 Reactome ORA
ereact <- enrichPathway(
  gene          = sig_genes,
  universe      = gene_universe,
  organism      = species_name,
  pAdjustMethod = "BH",
  readable      = TRUE
)

## 4.4 preranked GSEA (GO-BP)  -----------------------------------------------
# keep one row per ENTREZID – here we choose the row with the largest |logFC|
tbl_nodup <- tbl %>%                                # <- NEW
  group_by(ENTREZID) %>% 
  slice_max(order_by = abs(logFC), n = 1, with_ties = FALSE) %>% 
  ungroup()

ranks <- tbl_nodup %>%                              # <- use tbl_nodup
  arrange(desc(logFC)) %>% 
  transmute(ENTREZID, rnk = logFC) %>% 
  tibble::deframe()

gsea_go <- gseGO(
  geneList      = ranks,
  OrgDb         = orgDb,
  ont           = "BP",
  pAdjustMethod = "BH"
)

gsea_go_readable <- setReadable(
  gsea_go,          # numeric-ID version stays untouched
  orgDb, 
  keyType = "ENTREZID"
)


## ---- 4.4b  KEGG preranked GSEA --------------------------------------------
gsea_kegg <- gseKEGG(
  geneList      = ranks,        # already built for GO GSEA
  organism      = kegg_code,
  pAdjustMethod = "BH",
  verbose       = FALSE
)

# Save table
write.xlsx(as.data.frame(gsea_kegg),
           file = "output/KEGG_GSEA_results.xlsx", overwrite = TRUE)

# Extract neural-flavoured hits (same pattern as above)
gsea_kegg_df <- as.data.frame(gsea_kegg)
neural_kegg_gsea <- gsea_kegg_df %>% 
  dplyr::filter(grepl(neural_pattern, Description, ignore.case = TRUE))

cat("✓ KEGG GSEA: ", nrow(neural_kegg_gsea),
    " neural-keyword pathways (adj-P < 0.05)\n")

# Quick dot-plot (optional)
if (nrow(neural_kegg_gsea)) {
  make_dot(neural_kegg_gsea,
           "KEGG_GSEA_NEURAL",
           "Neural KEGG pathways (GSEA)")
}

## ---- 4.4c  Gene-level direction inside top GSEA sets -----------------------
topN_gsea <- 10                # adjust if you like
gsea_tbl  <- as.data.frame(gsea_go) %>% 
  slice_head(n = topN_gsea)

if (nrow(gsea_tbl)) {

  # Collapse leading-edge list ↴
  le_long <- gsea_tbl %>%
    tidyr::separate_rows(core_enrichment, sep = "/") %>%
    dplyr::rename(ENTREZID = core_enrichment) %>%
    left_join(tbl[, c("ENTREZID", "gene_label", "logFC")], by = "ENTREZID") %>%
    mutate(Direction = dplyr::case_when(
      logFC >  0   ~ "Up",
      logFC <  0   ~ "Down",
      TRUE         ~ "Neutral"
    ))

  # Counts per pathway
  dir_counts <- le_long %>%
    count(ID, Description, Direction, name = "Genes") %>%
    tidyr::pivot_wider(names_from = Direction, values_from = Genes,
                       values_fill = 0)

  print(dir_counts)  # also goes into the HTML/console

  # Stacked bar-plot
  library(ggplot2)
  p_dirs <- ggplot(le_long,
                   aes(x = reorder(Description, NES),
                       fill = Direction)) +
    geom_bar() +
    coord_flip() +
    labs(y = "Leading-edge genes",
         x = NULL,
         title = "Direction of leading-edge genes in top GSEA pathways") +
    scale_fill_manual(values = c(Up = "firebrick",
                                 Down = "forestgreen",
                                 Neutral = "grey70")) +
    theme_bw(base_size = 10)

  save_plot(p_dirs, "gsea_leading_edge_direction_bars.pdf",
            w = 8, h = 0.4 * topN_gsea + 2)

  # Export table for inspection
  openxlsx::write.xlsx(list(
    Top_GSEA            = gsea_tbl,
    LeadingEdge_Details = le_long,
    Direction_Counts    = dir_counts
  ),
  file = "output/gsea_leading_edge_direction.xlsx",
  overwrite = TRUE)
}

## 4.4d GO MF preranked GSEA  -----------------------------------------------
gsea_go_MF <- gseGO(
  geneList      = ranks,
  OrgDb         = orgDb,
  ont           = "MF",
  pAdjustMethod = "BH"
)

# Add to exports
write.xlsx(list(
  GO_BP_ORA     = as_df(ego),
  GO_MF_ORA     = as_df(ego_MF),
  KEGG_ORA      = as_df(ekegg),
  Reactome_ORA  = as_df(ereact),
  GO_BP_GSEA    = as_df(gsea_go,      "core_enrichment"),
  GO_MF_GSEA    = as_df(gsea_go_MF,   "core_enrichment"),
  Summary       = summary_stats
), file = "output/enrichment_results_with_genes_and_summary.xlsx", overwrite = TRUE)

## ---------------------------------------------------------------------------

## 5. EXPORT TABLES (OVERALL ANALYSIS) ----------------------------------------
as_df <- function(res, id_col = "geneID") {
  if (is.null(res) || nrow(as.data.frame(res)) == 0)
      return(data.frame())
  dplyr::rename(as.data.frame(res), Genes = !!id_col)
}

write.xlsx(list(
  GO_ORA   = as_df(ego),
  KEGG_ORA = as_df(ekegg),
  Reactome = as_df(ereact),
  GO_GSEA  = as_df(gsea_go, "core_enrichment"),
  Summary  = summary_stats
), file = "output/enrichment_results_with_genes_and_summary.xlsx",
   overwrite = TRUE)
write.xlsx(
  list(
    GO_ORA   = as_df(ego),                         # already SYMBOLs
    KEGG_ORA = as_df(ekegg),                       # already SYMBOLs
    Reactome = as_df(ereact),                      # already SYMBOLs
    GO_GSEA  = as_df(gsea_go_readable,             # ← NEW: readable copy
                     id_col = "core_enrichment"),
    Summary  = summary_stats
  ),
  file      = "output/enrichment_results_with_genes_and_summary.xlsx",
  overwrite = TRUE
)




## 6. BASIC VISUALISATIONS ----------------------------------------------------
# Volcano plot (all genes)
volc <- EnhancedVolcano(
  tbl,
  lab     = tbl$gene,
  x       = "logFC",
  y       = "padj",
  xlab    = bquote(~Log[2]~" fold change"),
  pCutoff = padj_cutoff,
  FCcutoff= logFC_cutoff,
  selectLab = tbl_sig$gene
)
save_plot(volc, "volcano_plot.pdf", 7, 6)

# Focused volcano (top N)
topNGenes <- 40
bigW <- 9; bigH <- 8
tbl_top <- tbl |>
  filter(padj < padj_cutoff) |>
  arrange(padj, desc(abs(logFC))) |>
  slice_head(n = topNGenes)

volc_focus <- EnhancedVolcano(
  tbl,
  lab        = tbl$gene,
  x          = "logFC",
  y          = "padj",
  pCutoff    = padj_cutoff,
  FCcutoff   = logFC_cutoff,
  selectLab  = tbl_top$gene,
  labSize    = 5,
  colAlpha   = .6,
  title      = glue("Focused volcano ({topNGenes} top genes highlighted)"),
  subtitle   = "Ordered by adjusted P-value then |log2FC|",
  caption    = ""
)
save_plot(volc_focus, "volcano_top_genes_large.pdf", bigW, bigH)

# logFC bins barplot
pdf("output/logFC_bins_barplot.pdf", 6, 4)
barplot(table(cut(abs(tbl$logFC),
                  c(0, 1, 2, 4, 8, Inf),
                  labels = c("<1", "1-2", "2-4", "4-8", ">8"))),
        main = "Distribution of |log2FC|",
        ylab = "Gene count",
        xlab = "|log2FC| bin")
dev.off()

## 7. UpSet enrichment plot ---------------------------------------------------
terms <- list()
if (nrow(as_df(ego))    > 0) terms$GO_BP    <- na.omit(unique(as_df(ego)$Description))
if (nrow(as_df(ekegg))  > 0) terms$KEGG     <- na.omit(unique(as_df(ekegg)$Description))
if (nrow(as_df(ereact)) > 0) terms$Reactome <- na.omit(unique(as_df(ereact)$Description))

if (length(terms) >= 2 && sum(lengths(terms) > 0) >= 2) {

## -- 8a: ComplexHeatmap :: UpSet (robust) ----------------------------------
comb_mat <- ComplexHeatmap::make_comb_mat(terms)

# Order combinations by size
comb_mat <- comb_mat[order(ComplexHeatmap::comb_size(comb_mat),
                           decreasing = TRUE)]

top_anno <- ComplexHeatmap::upset_top_annotation(
  comb_mat,
  gp = grid::gpar(fill = "grey60")   # ← no ‘border = …’ here
)

pdf("output/upset_enrichment_terms.pdf", 7, 5)
ComplexHeatmap::UpSet(
  comb_mat,
  top_annotation  = top_anno,
  right_annotation = NULL,
  show_row_names   = TRUE
)
dev.off()

} else {
  ## -- 7b: Fallback bar-plot -------------------------------------------------
  term_counts <- lengths(terms)
  bar_df <- data.frame(Category = names(term_counts),
                       Count    = as.numeric(term_counts))
  pdf("output/enrichment_term_counts_barplot.pdf", 6, 4)
  print(
    ggplot(bar_df, aes(x = Category, y = Count, fill = Category)) +
      geom_bar(stat = "identity") +
      theme_bw() +
      labs(title = "Enrichment term counts",
           y = "Number of terms", x = "Category")
  )
  dev.off()
}

## 8. Dot-plots and cnetplot ---------------------------------------------------
if (nrow(as_df(ego))) save_plot(cnetplot(ego, showCategory = 10,
                                         node_label = "all"),
                                "cnetplot_GO_ORA.pdf", 9, 8)
make_dot(ego,      "GO_ORA",      "GO BP ORA")
make_dot(ekegg,    "KEGG_ORA",    "KEGG ORA")
make_dot(ereact,   "Reactome_ORA","Reactome ORA")
make_dot(gsea_go,  "GO_GSEA",     "GO GSEA (preranked)")

## 9. EXTRA VISUALISATIONS ---------------------------------------------------
# 9.1 emapplot for GO ORA
if (nrow(as.data.frame(ego)) > 0) {
  ego_sim <- pairwise_termsim(ego)
  emapplot_out <- enrichplot::emapplot(ego_sim,
                                       showCategory = 30)
  save_plot(emapplot_out, "emapplot_GO_ORA.pdf", 9, 7)
}



# 9.2 GSEA running-score plots
if (nrow(as.data.frame(gsea_go)) > 0) {
  pdf("output/gsea_running_score.pdf", 9, 6)
  for (id in head(gsea_go@result$ID, 3))
      print(enrichplot::gseaplot2(gsea_go, geneSetID = id))
  dev.off()
}


## 10. Leading-edge export & barplots -----------------------------------------

if (nrow(as.data.frame(gsea_go)) > 0) {
  # pull out the top 3 GSEA terms
  top_gsea <- head(gsea_go@result, 3)
  
  # make sure we get dplyr::select, not the S4Generic select()
  leading_edge_long <- top_gsea |>
    dplyr::select(ID, Description, core_enrichment, NES, p.adjust) |>
    tidyr::separate_rows(core_enrichment, sep = "/") |>
    dplyr::rename(ENTREZID = core_enrichment)
  
  leading_edge_long <- leading_edge_long |>
    left_join(tbl[, c("ENTREZID", "gene", "logFC")], by = "ENTREZID")
  
  unmatched <- leading_edge_long %>% filter(is.na(logFC))
  if (nrow(unmatched) > 0) {
    warning("Some leading-edge genes could not be matched (no logFC):")
    print(unique(unmatched$ENTREZID))
  }

  write.xlsx(leading_edge_long,
             file = "output/gsea_leading_edge_genes.xlsx",
             overwrite = TRUE)

  for (term in unique(leading_edge_long$ID)) {
    le_term <- leading_edge_long[leading_edge_long$ID == term &
                                 !is.na(leading_edge_long$logFC), ]
    if (nrow(le_term) > 0) {
      topN_le <- le_term %>%
                 arrange(desc(abs(logFC))) %>%
                 head(25)
      p <- ggplot(topN_le,
                  aes(x = reorder(gene, logFC), y = logFC,
                      fill = logFC)) +
        geom_col() +
        coord_flip() +
        labs(title = glue("Leading-edge genes: {unique(topN_le$Description)}"),
             x = "Gene symbol", y = "log2 fold change") +
        theme_bw(base_size = 11) +
        scale_fill_gradient2(low = "royalblue", mid = "white",
                             high = "firebrick")
      save_plot(p,
                sprintf("gsea_leading_edge_%s.pdf", term), 8, 5)
    } else {
      message(glue("No matching logFC for leading-edge genes in term {term}."))
    }
  }
}

library(openxlsx) # if not already loaded

## ---- 10.b  Annotate KEGG ORA results with pathway CLASS --------------------
## Adds a `Class` column to ekegg@result and rebuilds the data.frame copy
if (nrow(as.data.frame(ekegg))) {

  message("Annotating KEGG pathways with CLASS (hierarchy) …")

  ekegg_df <- as.data.frame(ekegg)
  # Fetch CLASS strings for every pathway ID in one go
  kinfo <- KEGGREST::keggGet(ekegg_df$ID)

  class_df <- data.frame(
    ID    = vapply(kinfo, function(x) sub("path:", "", x$ENTRY),  ""),
    Class = vapply(kinfo, function(x)
                   paste(x$CLASS, collapse = "; "), ""),
    stringsAsFactors = FALSE
  )

  ekegg_df <- dplyr::left_join(ekegg_df, class_df, by = "ID")
  ekegg@result <- ekegg_df     # keep ekegg object in sync
}


## 11. KEGG PATHWAY DIAGRAMS + GUIDE ------------------------------------------
fc_vec <- setNames(tbl$logFC, tbl$ENTREZID)

kegg_outdir <- file.path("output", "pathview")
dir.create(kegg_outdir, showWarnings = FALSE, recursive = TRUE)

guide_list <- list()

if (nrow(as.data.frame(ekegg)) > 0) {
  kegg_df <- as.data.frame(ekegg)
  # Get up to N top pathways and their descriptions-> change this below if needed to see more 
  top_kegg <- head(kegg_df$ID, 10)
  top_desc <- head(kegg_df$Description, 10)

  for (i in seq_along(top_kegg)) {
    pid <- top_kegg[i]
    desc <- top_desc[i]
    clean_id <- sub("^gga", "", pid)
    message("Rendering KEGG diagram for ", pid)
    # Save old working directory
    old_wd <- getwd()
    setwd(kegg_outdir)
    tryCatch({
      pv_out <- pathview(
        gene.data  = fc_vec,
        pathway.id = clean_id,
        species    = kegg_code,
        out.suffix = "logFC"
      )
      # Pathview returns a list; get file base name
      # PNG and PDF filenames look like "gga04142.logFC.png", "gga04142.logFC.pdf"
      basefile <- paste0(pid, ".logFC")
      pngfile <- paste0(basefile, ".png")
      pdffile <- paste0(basefile, ".pdf")
      guide_list[[length(guide_list) + 1]] <- data.frame(
        Pathway_ID   = pid,
        Description  = desc,
        PNG_File     = file.path("pathview", pngfile),
        PDF_File     = file.path("pathview", pdffile),
        stringsAsFactors = FALSE
      )
    }, error = function(e) {
      message("Pathview error for ", pid, ": ", e$message)
    })
    setwd(old_wd)
  }
}

# Combine and write output guide
if (length(guide_list) > 0) {
  kegg_guide <- do.call(rbind, guide_list)
  write.csv(kegg_guide, file = file.path("output", "kegg_pathview_guide.csv"), row.names = FALSE)
  write.xlsx(kegg_guide, file = file.path("output", "kegg_pathview_guide.xlsx"), overwrite = TRUE)
  message("KEGG Pathview guide written to output/kegg_pathview_guide.csv and .xlsx")
}

## 12. NEURAL-FOCUSED SUB-ANALYSIS -------------------------------------------

pick_neural_terms <- function(df) {
  if (nrow(df) == 0 || !("Description" %in% colnames(df))) return(df)
  df[grepl(paste(neural_kw, collapse="|"), df$Description,
           ignore.case = TRUE), , drop = FALSE]
}

neural_ora   <- pick_neural_terms(as.data.frame(ego))
neural_gsea  <- pick_neural_terms(as.data.frame(gsea_go))

cat(sprintf("\nFound %d neural GO-BP terms in ORA; %d in preranked GSEA\n",
            nrow(neural_ora), nrow(neural_gsea)))

if (nrow(neural_ora)  > 0) print(neural_ora$Description)
if (nrow(neural_gsea) > 0) print(neural_gsea$Description)

## 13.1 Dot-plots (neural only) ----------------------------------------------
if (nrow(neural_ora)  > 0) make_dot(neural_ora,  "GO_ORA_NEURAL",
                                    "Neural GO-BP ORA (subset)")
if (nrow(neural_gsea) > 0) make_dot(neural_gsea, "GO_GSEA_NEURAL",
                                    "Neural GO-BP GSEA (subset)")

## 13.2 Ridge-plots -----------------------------------------------------------
# Ridgeplot helper vector
if (!exists("fc_vec"))
    fc_vec <- setNames(tbl$logFC, tbl$ENTREZID)

 #a dotplot for neural ORA terms
neural_ids  <- unique(na.omit(neural_ora$ID))
N_ora       <- min(length(neural_ids), 15)

if (N_ora > 0) {
  neural_ego        <- ego
  # Subset result to neural IDs only
  neural_ego@result <- dplyr::filter(ego@result, ID %in% neural_ids)
  # Plot: show all of them, or up to N_ora
  p <- enrichplot::dotplot(neural_ego, showCategory = N_ora,
                           orderBy = "GeneRatio") +
       labs(title = "Neural GO-BP ORA (subset)",
            x = "Gene ratio", y = NULL) +
       theme(axis.text.y = element_text(size = 10))
  pdf("output/dotplot_GO_ORA_NEURAL.pdf", 7, 0.5 * N_ora + 2)
  print(p); dev.off()
}

# GSEA ridgeplot
if (nrow(neural_gsea) >= 3 && "ID" %in% colnames(neural_gsea)) {
  neural_gsea_ids <- na.omit(neural_gsea$ID)
  neural_rows <- which(gsea_go@result$ID %in% neural_gsea_ids)
  if (length(neural_rows) >= 3) {
    N_gsea <- min(length(neural_rows), 15)
    plot_height <- 0.5 * N_gsea + 2
    gsea_go_neural <- gsea_go
    gsea_go_neural@result <- gsea_go@result[neural_rows, , drop = FALSE]
    pdf("output/ridgeplot_GO_GSEA_NEURAL.pdf", 7, plot_height)
    p <- enrichplot::ridgeplot(gsea_go_neural, showCategory = N_gsea) +
         labs(x = "Enrichment score (NES)", y = NULL) +
         theme(axis.text.y = element_text(size = 10))
    print(p); dev.off()
  }
}

## 13.3 Cnetplot (neural ORA) -------------------------------------------------
neural_ids  <- unique(na.omit(neural_ora$ID))
N_ora       <- min(length(neural_ids), 15)

if (N_ora > 0) {
  neural_ego        <- ego
  neural_ego@result <- dplyr::filter(ego@result, ID %in% neural_ids)

  # Only try if there are at least two terms
  if (nrow(neural_ego@result) >= 2) {
    p <- cnetplot(neural_ego,
                  showCategory = N_ora,
                  foldChange   = fc_vec,
                  node_label   = "all")
    save_plot(p, "cnetplot_GO_ORA_NEURAL.pdf", 10, 8)
  } else {
    message("Not enough neural GO terms for cnetplot (need at least 2). Skipping.")
  }
}

## 13.4 Heatmap of neural gene logFC -----------------------------------------
if ((nrow(neural_ora) > 0) | (nrow(neural_gsea) > 0)) {
  neural_gene_vec <- unique(unlist(strsplit(
    c(neural_ora$Genes,
      if ("core_enrichment" %in% colnames(neural_gsea))
          neural_gsea$core_enrichment else NULL), "/")))
  neural_gene_tbl <- tbl |> filter(ENTREZID %in% neural_gene_vec)
  if (nrow(neural_gene_tbl) > 0) {
    mat <- matrix(neural_gene_tbl$logFC, nrow = 1,
                  dimnames = list(NULL, neural_gene_tbl$gene))
    col_fun <- circlize::colorRamp2(
      c(-max(abs(mat)), 0, max(abs(mat))),
      c("royalblue", "white", "firebrick"))
    pdf("output/heatmap_neural_gene_logFC.pdf", 8, 2.5)
    ComplexHeatmap::Heatmap(mat, name = "log2FC", col = col_fun,
                            cluster_columns = TRUE,
                            show_row_names   = FALSE,
                            column_names_gp  = grid::gpar(fontsize = 7))
    dev.off()
  }
}

## 13.5 Save neural summary ---------------------------------------------------
neural_summary <- data.frame(
  Set            = c("Neural ORA", "Neural GSEA"),
  N_terms        = c(nrow(neural_ora), nrow(neural_gsea)),
  Top_term_ORA   = c(if (nrow(neural_ora)  > 0) neural_ora$Description[1] else NA, NA),
  Top_term_GSEA  = c(NA, if (nrow(neural_gsea) > 0) neural_gsea$Description[1] else NA)
)

write.xlsx(list(
  Neural_GO_ORA  = neural_ora,
  Neural_GO_GSEA = neural_gsea,
  Neural_Summary = neural_summary
), file = "output/enrichment_neural_only.xlsx", overwrite = TRUE)

library(ggplot2)
lead_df <- leading_edge_long   # already built in section 11

ggplot(lead_df, aes(x = reorder(gene, logFC), y = logFC)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Log₂FC of leading-edge neural genes",
       x = NULL, y = "log₂FC") +
  theme_bw(base_size = 10)

if (exists("leading_edge_long")) {
  plot_df <- leading_edge_long %>% filter(!is.na(logFC))
  
  cat("Rows with logFC:", nrow(plot_df), " / total:", nrow(leading_edge_long), "\n")
  
  if (nrow(plot_df) == 0) {
    warning("All leading-edge genes lack logFC values – check gene-ID mapping.")
  } else {
    library(ggplot2)
    ggplot(plot_df,
           aes(x = reorder(gene, logFC), y = logFC, fill = logFC)) +
      geom_col() +
      coord_flip() +
      scale_fill_gradient2(low = "royalblue", mid = "white", high = "firebrick") +
      labs(title = "Log₂FC of leading-edge neural genes",
           x = NULL, y = "log₂FC") +
      theme_bw(base_size = 10)
  }
} else {
  warning("Variable leading_edge_long is missing – was the GSEA block skipped?")
}


saveRDS(
  list(ego    = ego,
       ekegg  = ekegg,
       ereact = ereact,
       fc_vec = fc_vec),
  "output/enrichment_objects.rds"
)

## ---- 13.6  Pathview diagrams for *neural-signature* KEGG pathways ---------
## *Requires*: ekegg, fc_vec, kegg_code, neural_kw  (all defined earlier)

message("Searching KEGG ORA results for neural-related pathways …")

neural_kegg <- ekegg_df |>
  dplyr::filter(grepl(neural_pattern,
                      paste(Description, Class),  ## search BOTH fields
                      ignore.case = TRUE))

cat("✓ Found", nrow(neural_kegg), "neural KEGG pathways\n")

if (nrow(neural_kegg) > 0) {

  kegg_neural_dir <- file.path("output", "pathview_neural")
  dir.create(kegg_neural_dir, showWarnings = FALSE, recursive = TRUE)

  neural_guide <- list()

  for (i in seq_len(nrow(neural_kegg))) {
    pid   <- neural_kegg$ID[i]          # e.g. "gga04724"
    desc  <- neural_kegg$Description[i]
    clean <- sub("^gga", "", pid)       # numeric part for pathview()

    message(" → Rendering ", pid, "  (", desc, ")")

    old_wd <- getwd(); setwd(kegg_neural_dir)
    tryCatch({
      pv <- pathview(
        gene.data   = fc_vec,
        pathway.id  = clean,
        species     = kegg_code,
        out.suffix  = "logFC_neural",   # keeps files separate from earlier run
        kegg.native = TRUE
      )

      base <- paste0(pid, ".logFC_neural")
      neural_guide[[length(neural_guide) + 1]] <- data.frame(
        Pathway_ID  = pid,
        Description = desc,
        PNG_File    = file.path("pathview_neural", paste0(base, ".png")),
        PDF_File    = file.path("pathview_neural", paste0(base, ".pdf")),
        stringsAsFactors = FALSE
      )
    }, error = function(e) {
      message("   ⚠ Pathview error for ", pid, ": ", e$message)
    })
    setwd(old_wd)
  }

  if (length(neural_guide)) {
    guide_df <- do.call(rbind, neural_guide)
    write.csv( guide_df,
               file = file.path("output", "kegg_pathview_neural_guide.csv"),
               row.names = FALSE)
    openxlsx::write.xlsx(
      guide_df,
      file = file.path("output", "kegg_pathview_neural_guide.xlsx"),
      overwrite = TRUE
    )
    message("✓ Neural KEGG Pathview guide written to output/ (CSV & XLSX)")
  }

} else {
  message("No neural KEGG pathways detected – skipping neural pathview block.")
}

## ---- 13.7  Direction of leading-edge genes in neural GSEA ------------------
## Needs: neural_gsea, tbl, save_plot(), openxlsx already in memory

if (exists("neural_gsea") && nrow(neural_gsea)) {

  neural_gsea_df <- as.data.frame(neural_gsea)

  le_neural <- neural_gsea_df %>%                           # unpack leading edge
    tidyr::separate_rows(core_enrichment, sep = "/") %>%
    dplyr::rename(ENTREZID = core_enrichment) %>%
    dplyr::left_join(tbl[, c("ENTREZID", "gene_label", "logFC")],
                     by = "ENTREZID") %>%
    dplyr::mutate(Direction = dplyr::case_when(
      logFC >  0  ~ "Up",
      logFC <  0  ~ "Down",
      TRUE        ~ "Neutral"
    ))

  # Count Up / Down / Neutral per pathway
  dir_counts_neural <- le_neural %>%
    dplyr::count(ID, Description, Direction, name = "Genes") %>%
    tidyr::pivot_wider(names_from = Direction, values_from = Genes,
                       values_fill = 0)

  print(dir_counts_neural)    # appears in the knitted report

  ## Stacked bar-plot
  library(ggplot2)
  p_neural_dirs <- ggplot(le_neural,
                          aes(x = reorder(Description, NES), fill = Direction)) +
    geom_bar() +
    coord_flip() +
    labs(title = "Direction of leading-edge genes\n(neural GSEA pathways)",
         y = "Leading-edge genes", x = NULL) +
    scale_fill_manual(values = c(Up = "firebrick",
                                 Down = "forestgreen",
                                 Neutral = "grey70")) +
    theme_bw(base_size = 10)

  save_plot(p_neural_dirs,
            "gsea_neural_leading_edge_direction_bars.pdf",
            w = 8,
            h = 0.4 * n_distinct(le_neural$Description) + 2)

  ## Workbook
  openxlsx::write.xlsx(list(
    Neural_GSEA            = neural_gsea_df,
    LeadingEdge_Details    = le_neural,
    Direction_Counts       = dir_counts_neural
  ),
  file = "output/gsea_neural_leading_edge_direction.xlsx",
  overwrite = TRUE)

  message("✓ neural GSEA direction plot + workbook saved to output/")
} else {
  message("No neural GSEA pathways available – skipping direction analysis.")
}
## ---------------------------------------------------------------------------



## ----  EXTRA: logFC distribution panels  ------------------------------------
message("Building logFC distribution panels …")

# Re-build neural_gene_tbl if it isn’t in scope yet
if (!exists("neural_gene_tbl")) {
  neural_gene_vec <- unique(unlist(strsplit(
    c(if (exists("neural_ora"))  neural_ora$Genes else NULL,
      if (exists("neural_gsea") && "core_enrichment" %in% colnames(neural_gsea))
          neural_gsea$core_enrichment else NULL),
    "/")))
  neural_gene_tbl <- tbl |> filter(ENTREZID %in% neural_gene_vec)
}

# Assemble long table: all   vs significant vs neural
plot_df <- dplyr::bind_rows(
  tbl               |> dplyr::transmute(logFC, Category = "All mapped"),
  tbl_sig           |> dplyr::transmute(logFC, Category = "Significant"),
  neural_gene_tbl   |> dplyr::transmute(logFC, Category = "Neural")
)

# Order facets the way we like
plot_df$Category <- factor(plot_df$Category,
                           levels = c("All mapped", "Significant", "Neural"))

# Plot – histogram (density-scaled) + density curve
library(ggplot2)
p_lfc <- ggplot(plot_df, aes(x = logFC)) +
  geom_histogram(aes(y = ..density..),
                 bins = 50, fill = "grey75", colour = "black") +
  geom_density(colour = "darkblue", linewidth = 0.9, alpha = 0.6) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(~ Category, ncol = 1, scales = "free_y") +
  labs(title  = "Distribution of log₂ fold changes",
       x      = "log₂ fold change",
       y      = "Density") +
  theme_bw() +
  theme(strip.text = element_text(size = 11),
        plot.title = element_text(hjust = 0.5))

# Save via the helper you already defined
save_plot(p_lfc, "logFC_distribution_panels.pdf", w = 7, h = 9)

message("✓ logFC distribution panels saved to output/logFC_distribution_panels.pdf")






cat("\n✓ Neural-focused analysis complete – see 'output/'\n")

cat("\n✓ All done! – see 'output/'\n")

cat("Ready for the next step? Run the cnet visualization script or the direction specific script")




```



```{r compareCluster_up_down, message=FALSE}
###############################################################################
##  compareCluster (Up vs Down)  – quick & facet plots with GeneRatio        ##
###############################################################################

# Ensure output/extra exists -------------------------------------------------
dir.create(file.path("output", "extra"), showWarnings = FALSE)

# Only run if both gene lists are present ------------------------------------
if (exists("up_sig")   && nrow(up_sig)   > 0 &&
    exists("down_sig") && nrow(down_sig) > 0) {

  ## 1. Run compareCluster ----------------------------------------------------
  cmp <- compareCluster(
           geneCluster   = list(Up   = up_sig$ENTREZID,
                                Down = down_sig$ENTREZID),
           fun           = "enrichGO",
           OrgDb         = orgDb,
           ont           = "BP",
           universe      = gene_universe,
           pAdjustMethod = "BH",
           readable      = TRUE)

  ## 2. Quick single-panel dot-plot (optional) -------------------------------
  p_quick <- dotplot(cmp, showCategory = 15) +
             ggtitle("GO-BP compareCluster (Up vs Down)")
  save_plot(p_quick,
            file.path("extra", "compareCluster_GO_BP_UpDown_quick.pdf"),
            w = 10, h = 10)

  ## 3. Custom side-by-side facet plot with GeneRatio ------------------------
  # 3a. Tidy up the data frame
  cmp_df <- as.data.frame(cmp) %>%
            tidyr::separate(GeneRatio, c("g1","g2"), convert = TRUE) %>%
            dplyr::mutate(GeneRatioNum = g1 / g2)
  
  ## ──▼ NEW: keep only top N pathways per cluster ▼─────────────────────────
  N_SHOW <- 15           # ← change this number as you like
  cmp_df <- cmp_df %>%
            dplyr::group_by(Cluster) %>%
            dplyr::slice_min(p.adjust, n = N_SHOW)   # or slice_max(GeneRatioNum,…)

  # 3b. Shared y-axis ordering: largest GeneRatio across both clusters
  ordering <- cmp_df %>%
              dplyr::group_by(Description) %>%
              dplyr::summarise(maxGR = max(GeneRatioNum)) %>%
              dplyr::arrange(maxGR) %>%
              dplyr::pull(Description)
  cmp_df$Description <- factor(cmp_df$Description, levels = ordering)

  # 3c. Build the facet plot
  p_facets <- ggplot(cmp_df,
                     aes(x = GeneRatioNum,
                         y = Description,
                         colour = p.adjust,
                         size   = Count)) +
              geom_point() +
              facet_wrap(~ Cluster, nrow = 1, scales = "free_x") +
              scale_colour_continuous(low = "red", high = "blue",
                                      name = "adj-P") +
              scale_size(range = c(2, 6)) +
              labs(title = "GO-BP compareCluster (Up vs Down)",
                   x = "Gene ratio", y = NULL) +
              theme_bw() +
              theme(strip.text     = element_text(size = 12),
                    axis.text.y     = element_text(size = 9),
                    panel.spacing.x = unit(1.2, "lines"))

  save_plot(p_facets,
            file.path("extra",
                      "compareCluster_GO_BP_UpDown_GeneRatio_facets.pdf"),
            w = 10, h = 10)

    ## 3d.  Neural-only facet plot  --------------------------------------------
  cmp_df_neural <- cmp_df %>%
                   dplyr::filter(grepl(neural_pattern,
                                       Description,
                                       ignore.case = TRUE))

  if (nrow(cmp_df_neural) >= 3) {              # need at least a few terms

    # keep top N_SHOW per cluster (same rule as before)
    cmp_df_neural <- cmp_df_neural %>%
                     dplyr::group_by(Cluster) %>%
                     dplyr::slice_min(p.adjust, n = N_SHOW)

    # shared ordering for y-axis
    ordering_n <- cmp_df_neural %>%
                  dplyr::group_by(Description) %>%
                  dplyr::summarise(maxGR = max(GeneRatioNum)) %>%
                  dplyr::arrange(maxGR) %>%
                  dplyr::pull(Description)
    cmp_df_neural$Description <- factor(cmp_df_neural$Description,
                                        levels = ordering_n)

    p_neural <- ggplot(cmp_df_neural,
                       aes(x = GeneRatioNum,
                           y = Description,
                           colour = p.adjust,
                           size   = Count)) +
                geom_point() +
                facet_wrap(~ Cluster, nrow = 1, scales = "free_x") +
                scale_colour_continuous(low = "red", high = "blue",
                                        name = "adj-P") +
                scale_size(range = c(2, 6)) +
                labs(title = "Neural GO-BP compareCluster (Up vs Down)",
                     x = "Gene ratio", y = NULL) +
                theme_bw() +
                theme(strip.text     = element_text(size = 12),
                      axis.text.y     = element_text(size = 9),
                      panel.spacing.x = unit(1.2, "lines"))

    # dynamic height: ~0.25 in per term, min 4 in
    h_neural <- max(4, 0.25 * length(unique(cmp_df_neural$Description)) + 2)

    save_plot(p_neural,
              file.path("extra",
                        "compareCluster_GO_BP_UpDown_GeneRatio_neural_facets.pdf"),
              w = 9, h = h_neural)

    message("✓ neural compareCluster facet plot saved to output/extra/")

  } else {
    message("No neural-related GO terms found in compareCluster results.")
  }

  
  
  message("✓ compareCluster plots saved to output/extra/")

} else {
  message("Up- or Down-regulated gene lists missing – compareCluster skipped.")
}

```


**Notes**

* `KEGGREST` downloads the current *Gallus gallus* pathway mappings each run, so results stay up-to-date without the obsolete **KEGG.db** package.  
* When Pathview runs, the “gga” prefix is stripped (`clean_id <- sub("^gga", "", pid)`) because `pathview()` expects only the numeric code.  
* A single `fc_vec` object provides logFC values to Pathview **and** the neural ridgeplots, so everything stays in sync.

Run the notebook as usual; you should now see a non-empty **KEGG_ORA** sheet and an UpSet plot that includes KEGG terms.



